\textbf{Introducing the Choice-Confidence (CHOCO) Model for Bimodal Data from Subjective Ratings: Application to the Effect of Attractiveness on Reality Beliefs about AI-Generated Faces}

### 2025-06-14 paper/manuscript.tex
oki

\usepackage{longtable, booktabs, multirow, multicol, colortbl, hhline, caption, array, float, xpatch}
\usepackage{subcaption}
\renewcommand\thesubfigure{\Alph{subfigure}}
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.85}
\renewcommand{\textfraction}{0.15}
\renewcommand{\floatpagefraction}{0.7}

@@

\abstract{TO DO. }

\keywords{subjective scales, choice confidence model, reality beliefs,
AI-generated faces, attractiveness}

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Andy Field}{true}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, breakable, colback=white, opacityback=0, rightrule=.15mm, arc=.35mm, bottomrule=.15mm, left=2mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, leftrule=.75mm]

@@

\usepackage{pbalance} 
\usepackage{float}
\makeatletter
\let\oldtpt\ThreePartTable
\let\endoldtpt\endThreePartTable
\def\ThreePartTable{\@ifnextchar[\ThreePartTable@i \ThreePartTable@ii}
\def\ThreePartTable@i[#1]{\begin{figure}[!htbp]
\onecolumn
\begin{minipage}{0.5\textwidth}
\oldtpt[#1]
}
\def\ThreePartTable@ii{\begin{figure}[!htbp]
\onecolumn
\begin{minipage}{0.5\textwidth}
\oldtpt
}
\def\endThreePartTable{
\endoldtpt
\end{minipage}
\twocolumn
\end{figure}}
\makeatother

@@

Decision-making is often conceptualized as involving distinct processes:
the choice itself, and the confidence associated with that choice. In
experimental paradigms, these can be somewhat disentangled by prompting
participants to make a discrete choice selection (e.g., ``True''
vs.~``False''), followed by a separate confidence rating. However, this
artificial separation makes its joint analysis difficult, and may not
reflect real-world scenarios, where individuals often express both
choice and confidence simultaneously using a single, continuous scale.
In such scales, each side can represent a distinct latent category, and
the distance from the midpoint can indicate the level of confidence or
certainty. This integrated response format typically results in bimodal
distributions, with peaks corresponding to the mean confidence on either
side. Traditional beta regression models, which assume unimodal
distributions within the (0,1) interval, are ill-suited for such data.
One alternative is to transform the data into two variables a
posteriori: binarizing the side to represent choice and calculating the
absolute distance from the midpoint to represent confidence. These can
then be modeled separately, for instance, using logistic regressions for
choice and beta regressions for confidence (see
\citeproc{ref-makowski2025too}{Makowski, Te, et al., 2025} for an
example). While this approach can provide additional insights into
underlying mechanisms compared to a unique model, it assumes
psychological and statistical independence between choice and
confidence, which may not hold true in practice.

@@

Study 1 revealed that the Choice-Confidence (CHOCO) model was a much
better fit for bimodal bounded data, compared to other alternatives like
the Zero-and-One Inflated Beta (ZOIB) and Beta-Gate (Ordered Beta)
models. It also allowed for a deeper understanding through its
interpretable parameters, offering insights into possibly distinct
cognitive mechanisms, such as the probability of answering real
vs.~fake, and the associated confidence in these two choices. This was
illustrated by modelling the effect of sex on all the CHOCO parameters.

@@

Finally, we also show that the CHOCO model was able to capture reliable
and interpretable individual-level parameters, supporting its value to
measure inter-individual differences. An interesting dissociation
emerged between participant- and item-level variability: the latter
seemed mostly to be represented in the \emph{p} parameter, while
participants reliably varied in most of the components (aside from
\emph{bex}). This could suggest that external item characteristics
primarily influence the probability of being judged as real vs.~fake,
while the expressed confidence is first and foremost an individual
characteristic.

@@

The same analysis was run for Beauty, which indicated the following
differences: \textbf{ANA: can you report the main differences?}

@@

\textbf{WIP}

This study provides further evidence that attractive faces are more
likely to be judged as real (vs.~AI-generated), This effect was
particularly strong and robust among male participants. For females, the
effect was weaker in sample 1 but more pronounced in sample 2, which
might partly reflect increased statistical power. An alternative
possibility is that the second sample's experimental manipulation
introduced conflicting cues (e.g., faces labelled ``AI-generated'') that
could have disrupted intuitive heuristics, thereby increasing reliance
on attractiveness as a diagnostic cue, particularly in ambiguous trials.

@@

However, one might be concerned about the application of a continuous
distribution (the CHOCO mixture of two Beta distributions) to ordinal
data (in Sample 2, where a 7-point Likert scale was used), as it
translated effectively to only three unique values used to estimate a
Beta distribution. Ordinal ratings (limited number of discrete options
represented numerically by integers) are commonly used in psychology,
and often modeled as continuous. While this fact has spurred its own
polarized debate (REFS), our position is that the model used should
ideally reflect the data \emph{generating} mechanism rather than
necessarily focus on the \emph{observed} data\footnote{Ideally, the data
  recording method should be aligned with the assumed generating
  mechanism, which is an experiment design issue rather than a
  statistical one.} - although the former cannot always be easily
inferred or known. For the data of Sample 2, we held the assumption that
the underlying latent distribution was CHOCO distributed (as there is no
reason to assume it would be different from Sample 1), treated as
continuous CHOCO-distributed data, despite the fact that it was
collected on a 7-point scale. The motivation was to see if the model
could extrapolate and reliably estimate all the parameters based on a
limited number of data points but means in our case that, effectively,
each Beta distribution was extrapolated from the frequency of only three
unique values. While this is far from ideal, the CHOCO model still
proved insightful, although it might have impacted its ability detect
finer-grain changes, including quadratic relationships.

@@

\textbf{WIP}

This study introduces a new statistical model to analyze bimodal data,
which can be observed for instance when using subjective rating scales
in which the two sides correspond to a different ``choice'' (e.g.,
``false'' vs.~``true'', ``real'' vs ``fake''). The Choice-Confidence
(CHOCO) model conceptualizes the data as a mixture of two distributions,
one for each choice, and models the probability of choosing one or the
other as well as the degree of confidence in each choice. We validated
this model on real-life data by appling it to the reality beliefs of
Makowski, Te, et al. (\citeproc{ref-makowski2025too}{2025}), and showed
that it was able to capture the bimodal nature of the data and provides
a more accurate representation of the underlying processes than other
alternative approaches, including Zero-or-One inflated Beta (ZOIB)
models.

Little difference between Beauty and Attractiveness, aside that Beauty
revealed more effects in women. Why?

While our study was pertaining to \emph{simulation monitoring}, the
process of assessing and forming beliefs about the nature of external
stimuli (REF makowski phenomenological), we believe that it is part of
the mechanisms at play in the fabric of our sense of reality (REF
makowski thesis), alongside others, such as presence (the embodied
feeling of being physically ``in'' an experience) or \emph{reality
monitoring} (the process of distinguishing between internally generated
and externally perceived events, i.e., imagination vs.~perception).
Interestingly, recent studies about the latter support a possible
distinction at a neural level between \ldots{} and (binary) decision
making. (https://www.cell.com/neuron/fulltext/S0896-6273(25)00362-9).

Also: https://www.cell.com/neuron/fulltext/S0896-6273(16)30016-2

Future studies should assess psychometric property like the minimum
amount of data required to estimate and recover all parameters reliably.
As well as the minimum amount of unique value (for Likert scales)
requires to detect weak or non-linearly shaped effects.

Future studies should assess the ability of CHOCO models to also fit
unimodal distributions (so it can be used more generally), as well as
the benefits of more parsimonious parametrization (for instance, a
unique precision parameter controlling both the left and right sides, or
a linked confidence parameter e.g.~where \emph{confleft} is expressed as
a function of \emph{confright}).

@@

@@

@@

@@

>>>>

\usepackage{longtable, booktabs, multirow, multicol, colortbl, hhline, caption, array, float, xpatch}
\usepackage{subcaption}

\renewcommand\thesubfigure{\Alph{subfigure}}
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.85}
\renewcommand{\textfraction}{0.15}
\renewcommand{\floatpagefraction}{0.7}

@@

\abstract{As AI-generated content becomes increasingly indistinguishable
from real stimuli, understanding how individuals form beliefs about the
reality of what they perceive is both theoretically and practically
urgent. This study introduces the Choice-Confidence (CHOCO) model as a
tool to dissect the cognitive mechanisms underlying decision-making
judgments. Conceptualized as a mixture of two Beta distributions, CHOCO
simultaneously estimates the probability of selecting one category
(e.g., ``real'' vs.~``AI-generated'') and the confidence associated with
each choice. We apply this model to two datasets (N=141 and N=189) in
which participants judged whether faces were real photographs or
AI-generated images, finding that facial attractiveness systematically
increased the likelihood of being judged as ``real'', particularly for
male participants. We discuss how these findings relate to the broader
effect of facial attractiveness, and suggest future directions for both
the cognitive and psychometric application of the CHOCO framework. }

\keywords{subjective scales, choice confidence model, reality beliefs,
AI-generated faces, attractiveness, simulation monitoring}

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Andy Field}{true}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, breakable, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, left=2mm, rightrule=.15mm, arc=.35mm, opacityback=0, colback=white, leftrule=.75mm, toprule=.15mm]

@@

\usepackage{pbalance}
% \usepackage{float}
\makeatletter
\let\oldtpt\ThreePartTable
\let\endoldtpt\endThreePartTable
\def\ThreePartTable{\@ifnextchar[\ThreePartTable@i \ThreePartTable@ii}
\def\ThreePartTable@i[#1]{\begin{figure}[!htbp]
\onecolumn
\begin{minipage}{0.485\textwidth}
\oldtpt[#1]
}
\def\ThreePartTable@ii{\begin{figure}[!htbp]
\onecolumn
\begin{minipage}{0.48\textwidth}
\oldtpt
}
\def\endThreePartTable{
\endoldtpt
\end{minipage}
\twocolumn
\end{figure}}
\makeatother

@@

Decision-making is often conceptualized as involving distinct processes:
the choice itself, and the confidence associated with that choice
(\citeproc{ref-sanders2016signatures}{Sanders et al., 2016}). In
experimental paradigms, these can be somewhat disentangled by prompting
participants to make a discrete choice selection (e.g., ``True''
\emph{vs.} ``False''), followed by a separate confidence rating.
However, this artificial separation makes its joint analysis difficult,
and may not reflect real-world scenarios, where individuals often
express both choice and confidence simultaneously using a single,
continuous scale. In such scales, each side can represent a distinct
latent category, and the distance from the midpoint can indicate the
level of confidence or certainty. This integrated response format
typically results in bimodal distributions, with peaks corresponding to
the mean confidence on either side. Traditional beta regression models,
which assume unimodal distributions within the (0,1) interval, are
ill-suited for such data. One alternative is to transform the data into
two variables a posteriori: binarizing the side to represent choice and
calculating the absolute distance from the midpoint to represent
confidence. These can then be modeled separately, for instance, using
logistic regressions for choice and beta regressions for confidence (see
\citeproc{ref-makowski2025too}{Makowski, Te, et al., 2025} for an
example). While this approach can provide additional insights into
underlying mechanisms compared to a unique model, it assumes
psychological and statistical independence between choice and
confidence, which may not hold true in practice.

@@

Study 1 revealed that the Choice-Confidence (CHOCO) model was a much
better fit for bimodal bounded data, compared to other alternatives like
the Zero-and-One Inflated Beta (ZOIB) and Beta-Gate (Ordered Beta)
models. It also allowed for a deeper understanding through its
interpretable parameters, offering insights into possibly distinct
cognitive mechanisms, such as the probability of answering real
\emph{vs.} fake, and the associated confidence in these two choices.
This was illustrated by modelling the effect of sex on all the CHOCO
parameters.

@@

Finally, we also show that the CHOCO model was able to capture reliable
and interpretable individual-level parameters, supporting its value to
measure inter-individual differences. An interesting dissociation
emerged between participant- and item-level variability: the latter
seemed mostly to be represented in the \emph{p} parameter, while
participants reliably varied in most of the components (aside from
\emph{bex}). This could suggest that external item characteristics
primarily influence the probability of being judged as real \emph{vs.}
fake, while the expressed confidence is first and foremost an individual
characteristic.

@@

The same analysis was run for Beauty, which indicated the following
differences: \textbf{ANA: can you check and write down the main
differences?}

@@

This study provides further evidence that attractive faces are more
likely to be judged as real (\emph{vs.} AI-generated), This effect was
particularly strong and robust among male participants. For females, the
effect was weaker in sample 1 but more pronounced in sample 2, which
might partly reflect increased statistical power. An alternative
possibility is that the second sample's experimental manipulation
introduced conflicting cues (e.g., faces labelled ``AI-generated'') that
could have disrupted intuitive heuristics, thereby increasing reliance
on attractiveness as a diagnostic cue, particularly in ambiguous trials.

@@

However, one might be concerned about the application of the CHOCO model
(a mixture of two continuous Beta distributions) to ordinal data in
Sample 2 (where a 7-point Likert scale was used). Ordinal ratings
(limited number of discrete options represented numerically by integers)
are commonly used in psychology, and often modeled as continuous. While
this fact has spurred its own polarized debate
(\citeproc{ref-liddell2018analyzing}{Liddell \& Kruschke, 2018};
\citeproc{ref-norman2010likert}{Norman, 2010};
\citeproc{ref-sullivan2013analyzing}{Sullivan \& Artino Jr, 2013}), our
position is that the model used should ideally reflect the data
\emph{generating} mechanism rather than necessarily focus on the
\emph{observed} data\footnote{Ideally, the data recording method should
  be aligned with the assumed generating mechanism, which is an
  experiment design issue rather than a statistical one.} - although the
former cannot always be easily inferred or known. For the data of Sample
2, we held the assumption that the underlying latent distribution was
CHOCO distributed (as there is no reason to assume it would be different
from Sample 1), and thus treated as continuous CHOCO-distributed data,
despite the fact that it was collected on a 7-point scale. The
motivation was to see if the model could extrapolate and reliably
estimate all the parameters based on a limited number of data points but
means in our case that, effectively, each Beta distribution was
extrapolated from the frequency of only three unique values. While this
is far from ideal, the CHOCO model still proved insightful, although it
might have impacted its ability detect finer-grain changes, including
quadratic relationships.

@@

This study introduces a new statistical model to analyze bimodal data,
which can be observed when using subjective rating scales in which the
two sides correspond to a different category (e.g., ``false''
vs.~``true'', ``real'' vs ``fake'', ``positive'' vs.~``negative). The
Choice-Confidence (CHOCO) model conceptualizes the data as a mixture of
two distributions, one for each choice, and models the probability of
choosing one or the other as well as the degree of confidence in each
choice. We validated this model on real-life data by applying it to the
reality beliefs of Makowski, Te, et al.
(\citeproc{ref-makowski2025too}{2025}) as well as to a new sample,
showing that it captures the underlying cognitive processes more
accurately than alternative models (e.g., linear or ZOIB models), and
provides interpretable psychological parameters. These parameters, such
as the choice probability and respective confidence, offer a cognitively
meaningful decomposition of subjective decisions, applicable to a wide
range of tasks.

Beyond the statistical contribution, the study also replicates and
expands findings on the mechanisms that underpin simulation monitoring,
or judgments pertaining to beliefs about the real \emph{vs.} simulated
(or synthetic) nature of external stimuli. This process, important for
navigating and making sense of our environment, is likely contributing
to our sense of reality (the feeling and belief of being real in a real
environment, \citeproc{ref-makowski2018cognitive}{Makowski, 2018}),
together with other mechanisms such as presence (the embodied feeling of
being physically ``in'' an experience) and reality monitoring (the
process of distinguishing between internally generated and externally
perceived events, i.e., imagination \emph{vs.} perception).

Interestingly, recent studies about the latter support a possible
distinction at a neural level between a graded ``reality signal''
strength, primarily tracked by the fusiform gyrus, and its categorical
thresholding supporting the binary classification of reality \emph{vs.}
imagination, supported by a frontal network of brain regions, including
dorsomedial prefrontal cortex and the anterior insula
(\citeproc{ref-dijkstra2025neural}{Dijkstra et al., 2025}). Although
tangential, these findings might provide potential neuroscientific
empirical grounding for the CHOCO's model usefulness in investigating
dimensions of the sense of reality, opening the doors for future
research exploring the neural correlates of the CHOCO parameters.

The main empirical finding replicated in our study is that perceived
facial attractiveness predicts the likelihood of believing that a face
is real (\emph{vs.} AI-generated). In both samples, males were more
likely to identify highly attractive faces as real. The effect was
weaker for females, possibly clouded by statistical power and/or the
presence of other interacting moderators. Our work highlights how this
key social heuristic not only biases trait perceptions (i.e., ratings of
trustworthiness, competence, etc.) but also extends into other types of
judgments, including reality beliefs.

Despite these promising findings, several limitations must be
acknowledged. First, the stimuli shown to male and female participants
were not identical, each group rated faces of the opposite sex, limiting
our ability to make direct sex-based comparisons and fully answer the
potential sexual dimorphism existing in the formation of reality
beliefs. Second, all faces presented were real photographs, which could
raise concerns about the validity of the procedure. This however was a
deliberate design feature: our goal was not to assess the detectability
of current AI-generated images and study true discrimination abilities,
but rather to study the cognitive mechanisms that shape reality
judgments assuming the simulation is perceptually flawless.
Nevertheless, future work could leverage generative AI to systematically
manipulate facial attributes, such as attractiveness, while holding
other features constant, offering greater experimental control and the
ability to test causal hypotheses about the role of attractiveness for
simulation monitoring.

While our application focused on facial judgments, the CHOCO model has
broader potential. Many real-world rating tasks - whether about
emotions, morality, authenticity, political veracity, or aesthetic
appeal---produce bimodal or skewed response distributions that
traditional models fail to capture well. CHOCO offers a flexible,
interpretable approach that accommodates both the binary outcome of
decision-making and its related continuous evaluations. However, some
methodological and practical questions remain to be answered.

First, future studies should investigate the psychometric quality of the
CHOCO model to derive practical guidelines for its optimal use. These
include assessing the minimum amount of data required to reliably
estimate and recover the different CHOCO parameters. When applying the
model to ordinal data (such as Likert scales), another key question is
the minimum number of unique response categories needed to detect subtle
or nonlinear effects, such as quadratic modulation of confidence.
Second, assessing the model's robustness when applied to unimodal or
skewed data would extend its utility beyond strictly bimodal contexts
and support its general applicability. Third, more parsimonious variants
of the model could be explored, such as versions with a single shared
precision parameter for both sides or formulations where the confidence
on one side is modeled as a function of the other. Such simplifications
may retain interpretability while reducing parameter overhead and
improving model stability.

In conclusion, we developed and validated the CHOCO model, a
theoretically grounded and practically flexible tool for modeling
subjective judgments and delineating choice from confidence processes.
It not only advances statistical modeling of decision-making data but
opens new avenues for dissecting the cognitive mechanisms of belief,
confidence, and reality perception.

@@

\bibitem[\citeproctext]{ref-dijkstra2025neural}
Dijkstra, N., Rein, T. von, Kok, P., \& Fleming, S. M. (2025). A neural
basis for distinguishing imagination from reality. \emph{Neuron}.

@@

\bibitem[\citeproctext]{ref-liddell2018analyzing}
Liddell, T. M., \& Kruschke, J. K. (2018). Analyzing ordinal data with
metric models: What could possibly go wrong? \emph{Journal of
Experimental Social Psychology}, \emph{79}, 328--348.

@@

\bibitem[\citeproctext]{ref-norman2010likert}
Norman, G. (2010). Likert scales, levels of measurement and the
{``laws''} of statistics. \emph{Advances in Health Sciences Education},
\emph{15}, 625--632.

@@

\bibitem[\citeproctext]{ref-sanders2016signatures}
Sanders, J. I., Hangya, B., \& Kepecs, A. (2016). Signatures of a
statistical computation in the human sense of confidence. \emph{Neuron},
\emph{90}(3), 499--506.

\bibitem[\citeproctext]{ref-sullivan2013analyzing}
Sullivan, G. M., \& Artino Jr, A. R. (2013). Analyzing and interpreting
data from likert-type scales. \emph{Journal of Graduate Medical
Education}, \emph{5}(4), 541.
